Stool Microbiome Analysis

### 2021-11-25 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update analysis.

@@

```{r Supervised Filtering 5%}
# Define prevalence threshold as 5% of total samples
prevalenceThreshold <- 0.05 * nsamples(PSOtmao)
cat("The 10% prevalence threshold is", prevalenceThreshold, "samples.", "\n")

@@

Great. Do your due diligence and inspect each phyloseq object to see how the filtering and glomming steps changed the nubmer of taxonomy.

@@

# Family
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Family)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Family) +
  theme(legend.position="none")

# Genus - have to blow it up to make it useful
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Genus)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) +
  theme(legend.position="none")
```

@@

> # Test the impact of fiber cluster, accounting for covariates, using the log ratio test
dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
>
> # This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero. 
dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
>
> # The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
resultsNames(dds.ratio.batcho)
res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))

## Factored by TMAO Median 
```{r DESeq2 analysis - TMAO median}
# Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
sd <- sample_data(PSOtmao1G)
str(sd$sex)
str(sd$bmi_final.x)
# Factor sex, BMI okay
sample_data(PSOtmao1G)$sex = factor(sample_data(PSOtmao1G)$sex)

#~~~~~~~~~~~~~~~
# DESeq analysis
# TMAO median
#~~~~~~~~~~~~~~~
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_mdn)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio)
alpha = 0.05
alpha.big = 0.5
sigtab = res.ratio[which(res.ratio$padj < alpha.big), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

@@

```{r DESeq2 analysis - TMAO tertile}

# DESeq analysis - tmao_tertile
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_tertile)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
alpha = 0.05
sigtab = res.ratio[which(res.ratio$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

@@

@@

```{r Beta diversity calculations}
# PCoA
PCoA_WU <- ordinate(PSOtmao4G, "PCoA", "wunifrac")
PCoA_UnWU <- ordinate(PSOtmao4G, "PCoA", "uunifrac")
PCoA_BC <- ordinate(PSOtmao4G, "PCoA", "bray")

# Calculate distance matrix
dismax_wu = phyloseq::distance(PSOtmao4G, method="wunifrac", type = "samples")
dismax_uwu = phyloseq::distance(PSOtmao4G, method="unifrac", type = "samples")
dismax_bc = phyloseq::distance(PSOtmao4G, method="bray", type = "samples")

@@

```{r Beta diversity stats}
# Get sample_data
df <- as(sample_data(PSOtmao4G), "data.frame")
df$tmao_mdn <- relevel(df$tmao_mdn, ref = "below")

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_wu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_uwu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "unifrac", perm=999) # .001

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_bc ~ tmao_log + sex:age + bmi_final.y, data=df, method = "bray", perm=999) # .017

@@

Repeat looking at tertile 1 and 3. This section isn't currently working but I'm not sure why.

```{r ANCOM - genus level with covariates tertile}
source("../../Rscripts/ancom_v2.1.R")

# Step 2.1: Tertile 1 and 3
tert1_3 <- df2[df2$tmao_tertile == "Tertile1" | df2$tmao_tertile == "Tertile3",]
feature_table = feature_table[,colnames(feature_table) %in% rownames(tert1_3)]
meta_data = meta_data[rownames(meta_data) %in% rownames(tert1_3),]

# Define variables for ANCOM
main_var = "tmao_tertile"; p_adj_method = "BH"; alpha = 0.05
adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
# Run ANCCOM
t_start = Sys.time()
res = ANCOM(feature_table = feature_table, 
            meta_data = meta_data, 
            struc_zero = struc_zero, 
            main_var = main_var, 
            p_adj_method = p_adj_method, 
            alpha = alpha, 
            adj_formula = adj_formula, 
            rand_formula = rand_formula)
t_end = Sys.time()
t_run = t_end - t_start 
t_run # around 5s

# Merge with the phenotype data 
MetaData3 <- MetaData2 %>% dplyr::select(tmao_tertile)

# Gather the data
library(tidyr)
ggData2 <- gather(ggData,
                  key = "Taxa",
                  value = "Abundance",
                  -tmao_tertile)

# Plot it
p <- ggplot(data = ggData2,
            aes(x = Taxa, 
                y = Abundance,
                fill = tmao_tertile), 
            alpha = 0.1) +
  geom_boxplot(lwd=.25) +
  xlab(NULL) +
  ylab("Log of Taxa Abundance") +
  theme_bw() + # note order matters here - need bw to be before theme()
  theme(axis.text.y = element_text( hjust = 1),
        axis.text.x = element_text(angle = 60, hjust = 1),
        text=element_text(size=11))  +
  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
  #scale_fill_viridis(discrete = TRUE) +
  #coord_flip() +
  labs(fill = "TMAO \n Classification")

>>>>

# Factor sex, BMI okay
sample_data(PSOtmao)$sex = factor(sample_data(PSOtmao)$sex)

@@

```{r Supervised Filtering 0.05}
# Define prevalence threshold as 5% of total samples
prevalenceThreshold <- 0.05 * nsamples(PSOtmao)
cat("The 10% prevalence threshold is", prevalenceThreshold, "samples.", "\n")

@@

Great. Do your due diligence and inspect each phyloseq object to see how the filtering and glomming steps changed the number of taxonomy.

@@

# Family
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Family)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Family) +
  theme(legend.position="none")

# Genus - have to blow it up to make it useful
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Genus)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) +
  theme(legend.position="none")
```

@@

> DESeq2 Set Up
> Test the impact of fiber cluster, accounting for covariates, using the log ratio test
>
> dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
> This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero. 
>
> dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
> 
> The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
> 
> resultsNames(dds.ratio.batcho)
> res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))

## Factored by TMAO Median

```{r DESeq2 analysis - TMAO median}
#~~~~~~~~~~~~~~~
# DESeq analysis
# TMAO median
#~~~~~~~~~~~~~~~
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_mdn)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio)
alpha = 0.05
alpha.big = 0.5
sigtab = res.ratio[which(res.ratio$padj < alpha.big), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

@@

```{r DESeq2 analysis - TMAO tertile}
# DESeq analysis - tmao_tertile
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_tertile)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
alpha = 0.05
sigtab = res.ratio[which(res.ratio$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

@@

#~~~~~~~~~~~~~~~
# Plot Results Fold Change
#~~~~~~~~~~~~~~~
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  ylab("Taxonomy") +
  xlab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```
Save

```{r save deseq2 tmao tertile results}
write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv")
```

@@

```{r Beta diversity calculations}
# PCoA
PCoA_WU <- ordinate(PSOtmao1G, "PCoA", "wunifrac")
PCoA_UnWU <- ordinate(PSOtmao1G, "PCoA", "uunifrac")
PCoA_BC <- ordinate(PSOtmao1G, "PCoA", "bray")

# Calculate distance matrix
dismax_wu = phyloseq::distance(PSOtmao1G, method="wunifrac", type = "samples")
dismax_uwu = phyloseq::distance(PSOtmao1G, method="unifrac", type = "samples")
dismax_bc = phyloseq::distance(PSOtmao1G, method="bray", type = "samples")

@@

```{r save beta matrices}
save(PCoA_WU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
save(PCoA_UnWU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
save(PCoA_BC, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")

save(dismax_wu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
save(dismax_uwu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
save(dismax_bc, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
```

```{r Beta diversity stats}
# Get sample_data
df <- as(sample_data(PSOtmao1G), "data.frame")
df$tmao_mdn <- relevel(df$tmao_mdn, ref = "below")

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_wu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025
#adonis2(tmao_log ~ dismax_wu + sex*age + cystatinc_bd1, data=df, method = "wunifrac", perm=999)

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_uwu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "unifrac", perm=999) # .001
#adonis2(tmao_log ~ dismax_uwu + sex*age + cystatinc_bd1, data=df, method = "unifrac", perm=999)

@@

# ADONIS (perMANOVA) analysis
adonis2(dismax_bc ~ tmao_log + sex + age + bmi_final.y, data=df, method = "bray", perm=999) # .017
#adonis2(tmao_log ~ dismax_bc + sex*age + cystatinc_bd1, data=df, method = "bray", perm=999)

@@

p
```
# Stats for paper

Take a TMAO centric view point. So for the following questions, consider the whole group and the TMAO quantiles/tertiles.

```{r Desc.Stats set up}
# Get sample data groupings
sampleData <- as.data.frame(sample_data(PSOtmao1))
sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))

# Make lists of participants in each group
tertile1 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile1",])
tertile2 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile2",])
tertile3 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile3",])

# Get OTU table
otuTbl_PSO1 <- as.data.frame(otu_table(PSOtmao1))
otuTbl_PSO1 <- as.data.frame(t(otuTbl_PSO1)) # subjects to rows

# Get tax table
taxaData <- as.data.frame(tax_table(PSOtmao1))
taxaData$PCOFG <- paste0(taxaData$Phylum, taxaData$Class, taxaData$Order, taxaData$Family, taxaData$Genus)

How many OTUs (not glommed) in analysis?

```{r How many}
#~~~~~~~~~~
# Tertile 1
#~~~~~~~~~~
# What is the average number of OTUs per participant in tertile 1?
taxa_occurrences_T1 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile1,], function(x) sum(x>0))
summary(taxa_occurrences_T1)

otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
ppl_occurrences_T1 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile1,], function(x) sum(x>0))
summary(ppl_occurrences_T1)

#~~~~~~~~~~
# Tertile 2
#~~~~~~~~~~
# What is the average number of OTUs per participant in tertile 1?
taxa_occurrences_T2 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile2,], function(x) sum(x>0))
summary(taxa_occurrences_T2)

otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
ppl_occurrences_T2 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile2,], function(x) sum(x>0))
summary(ppl_occurrences_T2)

#~~~~~~~~~~
# Tertile 3
#~~~~~~~~~~
# What is the average number of OTUs per participant in tertile 1?
taxa_occurrences_T3 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile3,], function(x) sum(x>0))
summary(taxa_occurrences_T3)

otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
ppl_occurrences_T3 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile3,], function(x) sum(x>0))
summary(ppl_occurrences_T3)
```

How many OTUs are present in 5%, 25%, 50% of samples?

```{r}

What is the firmicutes to bacteroidetes ratio per tertile?
How common are Firmicutes? Mean, SD, range?
How common are Bacteroidetes? Mean, SD, range?

```{r Firmicutes to bacteroidetes ratio per tertile}
PSOtmao1_Phylum <- tax_glom(PSOtmao1, "Phylum", NArm = TRUE)
PSOtmao1_Phylum_tertile1 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile1")
PSOtmao1_Phylum_tertile2 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile2")
PSOtmao1_Phylum_tertile3 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile3")

#~~~~~~~~~~
# Tertile 1
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_Phy_tert1 = apply(X = otu_table(PSOtmao1_Phylum_tertile1),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_Phy_tert1 = data.frame(Prevalence = prevalence_Phy_tert1,
                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile1), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1_Phylum_tertile1))

# Firmicutes:Bacteroidetes Abundance Ratio
firm_bacter_t1 <- prevalence_Phy_tert1$TotalAbundance[prevalence_Phy_tert1$Phylum == " p__Firmicutes"] / prevalence_Phy_tert1$TotalAbundance[prevalence_Phy_tert1$Phylum == " p__Bacteroidetes"]

#~~~~~~~~~~
# Tertile 2
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_Phy_tert2 = apply(X = otu_table(PSOtmao1_Phylum_tertile2),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile2), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_Phy_tert2 = data.frame(Prevalence = prevalence_Phy_tert2,
                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile2), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1_Phylum_tertile2))

firm_bacter_t2 <- prevalence_Phy_tert2$TotalAbundance[prevalence_Phy_tert2$Phylum == " p__Firmicutes"] / prevalence_Phy_tert2$TotalAbundance[prevalence_Phy_tert2$Phylum == " p__Bacteroidetes"]

#~~~~~~~~~~
# Tertile 3
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_Phy_tert3 = apply(X = otu_table(PSOtmao1_Phylum_tertile3),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile3), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_Phy_tert3 = data.frame(Prevalence = prevalence_Phy_tert3,
                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile3), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1_Phylum_tertile3))

firm_bacter_t3 <- prevalence_Phy_tert3$TotalAbundance[prevalence_Phy_tert3$Phylum == " p__Firmicutes"] / prevalence_Phy_tert3$TotalAbundance[prevalence_Phy_tert3$Phylum == " p__Bacteroidetes"]

# Compare ratios
cat("Firmicutes:Bacteroidetes ratio tertile 1:", firm_bacter_t1)
cat("Firmicutes:Bacteroidetes ratio tertile 2:", firm_bacter_t2)
cat("Firmicutes:Bacteroidetes ratio tertile 3:", firm_bacter_t3)

What is the % abundance of a given taxa? Get summary statistics per tertile.

Range percentage of taxa per person = n occurrences of bug/total occurrences of all bugs * 100

```{r Firmicute summary}
#~~~~~~~~~~
# Tertile 1
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile1, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Firmicutes", ]
cat("Summmary of Firmicute abundance in tertile 1:")
summary(subdf$Abundance)

#~~~~~~~~~~
# Tertile 2
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile2, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Firmicutes", ]
cat("Summmary of Firmicute abundance in tertile 2:")
summary(subdf$Abundance)

#~~~~~~~~~~
# Tertile 3
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile3, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Firmicutes", ]
cat("Summmary of Firmicute abundance in tertile 3:")
summary(subdf$Abundance)

```{r Bacteroidetes summary}
#~~~~~~~~~~
# Tertile 1
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile1, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
cat("Summmary of Bacteroidetes abundance in tertile 1:")
summary(subdf$Abundance)

#~~~~~~~~~~
# Tertile 2
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile2, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
cat("Summmary of Bacteroidetes abundance in tertile 2:")
summary(subdf$Abundance)

#~~~~~~~~~~
# Tertile 3
#~~~~~~~~~~
trans <- transform_sample_counts(PSOtmao1_Phylum_tertile3, function(x) x / sum(x))
melted <- psmelt(trans)
subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
cat("Summmary of Bacteroidetes abundance in tertile 3:")
summary(subdf$Abundance)

Genus: What are the most common genus?
Use the PSOtmao1G, subset to low-tertile, make prevalence data frame, sort in descending order. Repeat for other tertiles.

```{r Genus Prevalence by TMAO Tertile}
# Subset by tertiles 
PSOtmao1_tertile1 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile1")
PSOtmao1_tertile2 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile2")
PSOtmao1_tertile3 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile3")

# How many taxa are there per tertile?
cat("n taxa in tertile 1:", ntaxa(PSOtmao1_tertile1))
cat("n taxa in tertile 2:", ntaxa(PSOtmao1_tertile2))
cat("n taxa in tertile 3:", ntaxa(PSOtmao1_tertile3))

# Now subset at the genus level to identify the most common genus
# Subset by tertiles (Genus)
PSOtmao1G_tertile1 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile1")
PSOtmao1G_tertile2 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile2")
PSOtmao1G_tertile3 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile3")

#~~~~~~~~~~
# Tertile 1
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_tert1 = apply(X = otu_table(PSOtmao1G_tertile1),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_tert1 = data.frame(Prevalence = prevalence_tert1,
                    TotalAbundance = taxa_sums(PSOtmao1G_tertile1), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1G_tertile1))

#~~~~~~~~~~
# Tertile 2
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_tert2 = apply(X = otu_table(PSOtmao1G_tertile2),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile2), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_tert2 = data.frame(Prevalence = prevalence_tert2,
                    TotalAbundance = taxa_sums(PSOtmao1G_tertile2), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1G_tertile2))
#~~~~~~~~~~
# Tertile 3
#~~~~~~~~~~
# Prevalence - how many samples the taxa is observed in
prevalence_tert3 = apply(X = otu_table(PSOtmao1G_tertile3),
               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile3), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevalence_tert3 = data.frame(Prevalence = prevalence_tert3,
                    TotalAbundance = taxa_sums(PSOtmao1G_tertile3), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao1G_tertile3))
```

View and sort top 10
```{r View abundant genus top 10}
cat("Most abundant genus in T1:")
head(prevalence_tert1[order(prevalence_tert1$TotalAbundance, decreasing= TRUE),], n = 10)
cat("Most abundant genus in T2:")
head(prevalence_tert2[order(prevalence_tert2$TotalAbundance, decreasing= TRUE),], n = 10)
cat("Most abundant genus in T3:")
head(prevalence_tert3[order(prevalence_tert3$TotalAbundance, decreasing= TRUE),], n = 10)

```{r Save tertile prevalence csv}
write.csv(prevalence_tert1, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile1.csv")
write.csv(prevalence_tert2, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile2.csv")
write.csv(prevalence_tert3, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile3.csv")
```

# Correlate Fecal Microbiome with Phenos

Now that we identified microbiome characteristics that are relevant to TMAO, ask if they

1) Improve the variance explained between TMAO and a given cardiometabolic marker
2) Correlate to cardiometabolic phenos

```{r Get the microbiome data ready to combine with pheno data}
da <- read.csv("../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv", row.names = 1)
resOTU <- rownames(da)
otuData <- as.data.frame(otu_table(PSOtmao1G))
otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
otuData_desRes <- as.data.frame(t(otuData_desRes))
# rename for ease downstream
names(otuData_desRes)[names(otuData_desRes) == "bb1b75f41ff9c9db1d1de41e8388eb52"] <- "f__Bacteroidaceae_g__Bacteroides"
names(otuData_desRes)[names(otuData_desRes) == "b180a2455b236c103cf0bfe620095736"] <- "f__Peptostreptococcaceae_g__"
names(otuData_desRes)[names(otuData_desRes) == "e1a2800b24cdf9779b28dc897cddb12a"] <- "f__Christensenellaceae_g__"
names(otuData_desRes)[names(otuData_desRes) == "8f5ceded1cd7e86b8271d3fab24d322a"] <- "f__Lachnospiraceae_g__Butyrivibrio"
```

Next chunk is borrowed from "Characteristics_Diet_Cardiometabolic"

```{r Load phenotype data}
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")

# Complete TMAO data only
phen <- phen[complete.cases(phen$tmao == TRUE),]

# Get rid of NA ethnicity category
phen$ethnicity <- droplevels(phen$ethnicity)

# Factor and recode sex
phen$sex <- as.factor(phen$sex)
phen$sex <- plyr::revalue(phen$sex, c("1" = "Male", "2" = "Female"))

# Make TMAO log variable
phen$tmao_log <- log(phen$tmao)

# Quantile
tmao_quantile <- quantile(phen$tmao)
phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))

# Check it
table(phen$tmao_quantile)
phen$tmao_quantile <- as.factor(phen$tmao_quantile)
str(phen$tmao_quantile)

# Tertile
tmao_tertile <- quantile(phen$tmao, c(0:3/3))
phen$tmao_tertile <- ifelse(phen$tmao <= tmao_tertile[2], "Tertile1",
                             ifelse(phen$tmao > tmao_tertile[2] & phen$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
# Factor it
phen$tmao_tertile <- factor(phen$tmao_tertile)
table(phen$tmao_tertile)

# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
phen$glc_bd1_ln <- log(phen$glc_bd1)
phen$tg_bd1_ln <- log(phen$tg_bd1)
phen$crp_bd1_ln <- log(phen$crp_bd1)
phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
phen$il6_bd1_ln <- log(phen$il6_bd1)
```

Merge phenotype and microbiome data for linear regression analysis
```{r}
phen <- merge(phen, otuData_desRes, by=0)
rownames(phen) <- phen[,"Row.names"]
phen <- subset(phen, select = -c(Row.names))
```

Run the regression!

f__Bacteroidaceae_g__Bacteroides + f__Peptostreptococcaceae_g__ + f__Christensenellaceae_g__ + f__Lachnospiraceae_g__Butyrivibrio

```{r Cardiometabolic + Microbiome Multiple Linear Regression}
# Transfored variable list
transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log", "f__Bacteroidaceae_g__Bacteroides", "f__Peptostreptococcaceae_g__", "f__Christensenellaceae_g__", "f__Lachnospiraceae_g__Butyrivibrio")

#~~~~~~~~
# TMAO  
#~~~~~~~~
# Make a smaller df from master (the old data)
phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Bacteroidaceae_g__Bacteroides, f__Peptostreptococcaceae_g__, f__Christensenellaceae_g__, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
n <- ncol(cvd_biomark)

# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1, phen_sub )) # include kcal
#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ ethnicity, phen_sub ))
#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ cystatinc_bd1, phen_sub ))

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
my_lm361_lcms_cvd <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm361_lcms_cvd) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm361_lcms_cvd) <- colnames(cvd_biomark)
res <- as.data.frame(my_lm361_lcms_cvd)

# BH correction
res$p.adj <- stats::p.adjust(res$P, method = "BH", n = n)
